---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

### *RGraphSpace*: Rendering Graphs as Coherent Spatial Objects in *ggplot2*

<!-- badges: start -->
  [![CRAN status](https://www.r-pkg.org/badges/version/RGraphSpace)](https://cran.r-project.org/package=RGraphSpace)
  [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
  [![CRAN downloads](https://cranlogs.r-pkg.org/badges/RGraphSpace)](https://cran.r-project.org/package=RGraphSpace)
  [![License: Artistic-2.0](https://img.shields.io/badge/license-Artistic--2.0-blue.svg)](https://cran.r-project.org/web/licenses/Artistic-2.0)
  [![DOI](https://img.shields.io/badge/doi-10.32614/CRAN.package.RGraphSpace-blue.svg)](https://doi.org/10.32614/CRAN.package.RGraphSpace)
  [![R-CMD-check](https://github.com/sysbiolab/RGraphSpace/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sysbiolab/RGraphSpace/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

### Highlights
* Native *ggplot2* interface for *igraph* objects
* Coherent rendering of node and edge layers
* Optimized *geoms* for high-dimensional data
* Spatial alignment to external reference frames

### Tutorials
* https://sysbiolab.github.io/RGraphSpace

### Overview

*RGraphSpace* is an R package that generates *ggplot2* graphics for *igraph* objects ([Csardi and Nepusz 2006](#ref-nepusz)) within a normalized coordinate space. The package implements new geometries that treat a graph as a single coherent object, synchronizing node and edge layers under standard aesthetic mappings. Node features are resolved on demand, supporting high-dimensional data without expanding node tables. Spatial alignment is available at the pixel level, with node coordinates anchored to pixel centers through a half-pixel offset, enabling precise node placement over external reference frames such as images and maps. The package also interoperates with *ggraph* and *tidygraph* workflows ([Pedersen 2024](#ref-pedersen)). Three specialized geoms translate graph data into geometric layers.

1. **`geom_nodespace()`**: Renders network nodes. Extends `GeomPoint`
aesthetic mappings and exposes node state information to the edge layer.

2. **`geom_edgespace()`**: Renders the relationships between nodes. Extends
`GeomSegment` aesthetic mappings; unlike standard segments, it is node-aware
and dynamically adjusts start and end points based on node position and size.

3. **`geom_graphspace()`**: A convenience wrapper that calls
`geom_nodespace()` and `geom_edgespace()` in sequence. Use this for the
common case; use the individual geoms directly when independent control of
node and edge layers is needed.

Core functionality relies only on lightweight R packages; optional geometry and large raster-image features use 'sf' and 'terra' (and their system libraries) when installed.

### Installation in R (>=4.5)

##### Release version

```r
# Release version from CRAN
install.packages("RGraphSpace")
```

##### Development version

```r
# Dependencies to build the vignettes
install.packages("knitr")
install.packages("rmarkdown")
install.packages("remotes")

# Package source
remotes::install_github("sysbiolab/RGraphSpace", build_vignettes=TRUE)
```

### Citation

If you use *RGraphSpace*, please cite:

```{r, echo=FALSE, results='asis'}
print(citation("RGraphSpace"), style = "text")
```

### Licenses

The *RGraphSpace* package is distributed under [Artistic-2.0](https://www.r-project.org/Licenses/Artistic-2.0)

### References

<a name="ref-wickham"></a>

* Wickham H: *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York, 2016. [https://ggplot2.tidyverse.org](https://ggplot2.tidyverse.org)

<a name="ref-nepusz"></a>

* Csardi G, Nepusz T: The igraph software package for complex network research. *InterJournal, Complex Systems* 1695. 2006. [https://igraph.org](https://igraph.org)

<a name="ref-pedersen"></a>

* Pedersen T: *tidygraph: A Tidy API for Graph Manipulation*. R package version 1.3.1. 2024. [doi:10.32614/CRAN.package.tidygraph](https://doi.org/10.32614/CRAN.package.tidygraph)

