---
title: "*RGraphSpace*: A lightweight interface between *igraph* and *ggplot2* graphics."
author: "Sysbiolab - Bioinformatics and Systems Biology Laboratory"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
abstract: "*RGraphSpace* is an interface to integrate *igraph* and *ggplot2* graphics in a normalized coordinate system. The package implements new geometric objects using *ggplot2* prototypes, customized for side-by-side visualization of multiple graphs. By scaling shapes and graph elements, *RGraphSpace* can provide a framework for layered visualizations."
output: 
  html_document:
    theme: cerulean
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    css: custom.css
vignette: >
  %\VignetteIndexEntry{"RGraphSpace: ggplot2 graphics for igraph"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br/>
**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

# Overview

*RGraphSpace* is an R package designed to generate *ggplot2* graphics for *igraph* objects [@Nepusz2006], scaling them to fit within a standardized unit space. This facilitates the side-by-side visualization of multiple graphs. The package implements new geometric objects based on *ggplot2* prototypes [@Wickham2016], accounting for the relative sizes of nodes and edges. By scaling graph elements, *RGraphSpace* supports layered visualizations with consistent spatial alignment. Its integration with *ggplot2* enables extensive customization of aesthetics and visual style, including colors, shapes, and line types.

# Quick start

This section will create a toy *igraph* object to demonstrate the *RGraphSpace* workflow. The graph layout is configured manually to ensure that users can easily view all the relevant arguments needed to prepare the input data for the *RGraphSpace* package. We will use the igraph's `make_star()` function to create a simple star-like graph and then the `V()` and `E()` functions to set attributes for vertices and edges, respectively. The *RGraphSpace* package will require that all vertices have `x`, `y`, and `name` attributes.

```{r Load packages for quick start, eval=TRUE, message=FALSE}
#--- Load required packages
library(igraph)
library(ggplot2)
library(RGraphSpace)
```

```{r Toy igraph - 1, eval=TRUE, message=FALSE, results=FALSE}
# Make a 'toy' igraph with 5 nodes and 4 edges;
# ..either a directed or undirected graph
gtoy1 <- make_star(5, mode="out")

# Check whether the graph is directed or not
is_directed(gtoy1)
## [1] TRUE

# Check graph size
vcount(gtoy1)
## [1] 5
ecount(gtoy1)
## [1] 4

# Assign 'x' and 'y' coordinates to each vertex;
# ..this can be an arbitrary unit in (-Inf, +Inf)
V(gtoy1)$x <- c(0, 2, -2, -4, -8)
V(gtoy1)$y <- c(0, 0,  2, -4,  0)

# Assign a name to each vertex
V(gtoy1)$name <- paste0("n", 1:5)
```

```{r Toy igraph - 2, eval=TRUE, message=FALSE, out.width="100%"}
# Plot the 'gtoy1' using standard R graphics
plot(gtoy1)
```

```{r Toy igraph - 3, eval=TRUE, message=FALSE, out.width="80%"}
# Plot the 'gtoy1' using RGraphSpace
plotGraphSpace(gtoy1, add.labels = TRUE)
```

# *RGraphSpace* attributes

Next, we will demonstrate all vertex and edge attributes that can be passed to *RGraphSpace* methods.

## Vertex attributes

```{r Node attributes, eval=TRUE, message=FALSE}
# Node size (numeric in (0, 100), as '%' of the plot space)
V(gtoy1)$nodeSize <- c(8, 5, 5, 5, 5)

# Node shape (integer code between 0 and 25; see 'help(points)')
V(gtoy1)$nodeShape <- c(21, 22, 23, 24, 25)

# Node color (Hexadecimal or color name)
V(gtoy1)$nodeColor <- c("red", "#00ad39", "grey80", "lightblue", "cyan")

# Node line width (as in 'lwd' standard graphics; see 'help(gpar)')
V(gtoy1)$nodeLineWidth <- 1

# Node line color (Hexadecimal or color name)
V(gtoy1)$nodeLineColor <- "grey20"

# Node labels ('NA' will omit labels)
V(gtoy1)$nodeLabel <- c("V1", "V2", "V3", "V4", NA)

# Node label size (in pts)
V(gtoy1)$nodeLabelSize <- 8

# Node label color (Hexadecimal or color name)
V(gtoy1)$nodeLabelColor <- "black"
```

## Edge attributes

Given a list of edges, *RGraphSpace* represents only one edge for each pair of connected vertices. If there are multiple edges connecting the same vertex pairs, it will display the line attributes of the first edge in the list.

```{r Edge attributes - 1, eval=TRUE, message=FALSE}
# Edge width (as in 'lwd' standard graphics; see 'help(gpar)')
E(gtoy1)$edgeLineWidth <- 0.8

# Edge color (Hexadecimal or color name)
E(gtoy1)$edgeLineColor <- c("red","green","blue","black")

# Edge type (as in 'lty' standard graphics; see 'help(gpar)')
E(gtoy1)$edgeLineType <- c("solid", "11", "dashed", "2124")

# Edge weight (numeric >=0; not passed to ggplot)
E(gtoy1)$weight <- 1
```

## Arrowhead attributes

Arrowhead in directed graphs: By default, an arrow will be drawn for each edge according to its left-to-right orientation in the edge list (*e.g.* `A -> B`).

```{r Edge attributes - 2, eval=TRUE, message=FALSE}
# Arrowhead types in directed graphs (integer code or character)
## 0 = "---", 1 = "-->", -1 = "--|"
E(gtoy1)$arrowType <- 1

# Arrowhead length (as in 'lwd' standard graphics; see 'help(gpar)')
E(gtoy1)$arrowLength <- 1
```

Arrowhead in undirected graphs: By default, no arrow will be drawn in undirected graphs.

```{r Edge attributes - 3, eval=TRUE, message=FALSE}
# Arrowhead types in undirected graphs (integer or character code)
##  0 = "---"
##  1 = "-->",  2 = "<--",  3 = "<->",  4 = "|->",
## -1 = "--|", -2 = "|--", -3 = "|-|", -4 = "<-|", 
E(gtoy1)$arrowType <- 1
# Note: in undirected graphs, this attribute overrides the 
# edge's orientation in the edge list

# Arrowhead length (as in 'lwd' standard graphics; see 'help(gpar)')
E(gtoy1)$arrowLength <- 1
```

... and now plot the updated *igraph* object with *RGraphSpace*:
 
```{r A shortcut for RGraphSpace, eval=TRUE, message=FALSE, out.width="80%"}
# Plot the updated 'gtoy1' using RGraphSpace
plotGraphSpace(gtoy1, add.labels = TRUE)
```

# Interactive layout

The following example demonstrates interoperability between *RGraphSpace* and *RedeR*, a package for interactive network visualization and manipulation.

```{r A shortcut for RedeR, eval=FALSE, message=FALSE}
# Load RedeR, a graph package for interactive visualization
## Note: this example requires Bioc >= 3.19
if(!require("BiocManager", quietly = TRUE)){
  install.packages("BiocManager")
  #BiocManager::install(version = "3.19")
}
if(!require("RedeR", quietly = TRUE)){
  BiocManager::install("RedeR")
}

# Launch the RedeR application
library(RedeR)
startRedeR()
resetRedeR()

# Send 'gtoy1' to the RedeR interface
addGraphToRedeR(gtoy1, unit="npc")
relaxRedeR()

# Fetch 'gtoy1' with a fresh layout
gtoy2 <- getGraphFromRedeR(unit="npc")

# Check the round trip...
plotGraphSpace(gtoy2, add.labels = TRUE)

## Note that for the round trip, shapes and line types are
## partially compatible between ggplot2 and RedeR.

# ...alternatively, just update the graph layout
gtoy2 <- updateLayoutFromRedeR(g=gtoy1)

# ...check the updated layout
plotGraphSpace(gtoy2, add.labels = TRUE)
```

# Other examples

The following vignettes illustrate how *RGraphSpace* can be used in combination with *PathwaySpace* to project network signals into landscape images.

[Projection of network signals](https://sysbiolab.github.io/PathwaySpace/){target="_blank" rel="noopener"}

# Citation

If you use *RGraphSpace*, please cite:

* Sysbiolab Team. "RGraphSpace: A lightweight package for representing large igraph objects in a normalized coordinate system." R package, 2023. Doi: 10.32614/CRAN.package.RGraphSpace

* Castro MA, Wang X, Fletcher MN, Meyer KB, Markowetz F (2012). "RedeR: R/Bioconductor package for representing modular structures, nested networks and multiple levels of hierarchical associations." *Genome Biology*, 13(4), R29. Doi: 10.1186/gb-2012-13-4-r29


# Session information
```{r label='Session information', eval=TRUE, echo=FALSE}
sessionInfo()
```


# References

