---
title: "Building a GraphSpace"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width="90%"
  )
```

<br/>
**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

```{r Check required version, eval=TRUE, message=FALSE}
# Check required version
if (packageVersion("RGraphSpace") < "1.5.0"){
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}
```

# Quick start

```{r Load packages, eval=TRUE, message=FALSE}
#--- Load required packages
library("RGraphSpace")
library("igraph")
library("ggplot2")
library("tidygraph")
```

This section creates a toy `igraph` from scratch to demonstrate the vertex and edge attributes that *RGraphSpace* parses automatically, showing exactly what the package expects as input. The same graph can also be supplied as a *tidygraph* object. Here we use `igraph`'s `make_star()` function and then `V()` and `E()` to assign attributes.

```{r Toy igraph - 1, eval=TRUE, message=TRUE}
# Make a 'toy' igraph with 5 nodes and 4 edges;
# ..either a directed or undirected graph
gtoy1 <- make_star(5, mode = "out")

# Check whether the graph is directed or not
is_directed(gtoy1)

# Check graph size
vcount(gtoy1)

ecount(gtoy1)

# Assign 'x' and 'y' coordinates to each vertex;
# ..this can be an arbitrary unit in (-Inf, +Inf)
V(gtoy1)$x <- c(0, 2, -2, -4, -8)
V(gtoy1)$y <- c(0, 0,  2, -4,  0)

# Assign a name to each vertex
V(gtoy1)$name <- paste0("n", 1:5)
```

```{r Toy igraph - 2, eval=TRUE, message=FALSE}
# The most direct call: pass an igraph to plotGraphSpace()
plotGraphSpace(gtoy1, node.labels = TRUE)
```

The same graph can be supplied as a *tidygraph* object; every *RGraphSpace* entry point accepts it through the same interface.

```{r Toy igraph - 3, eval=TRUE, message=FALSE}
# Same toy graph, as tidygraph
gr <- as_tbl_graph(gtoy1)
gr
```

```{r Toy igraph - 4, eval=TRUE, message=FALSE}
plotGraphSpace(gr, node.labels = TRUE)
```


# *RGraphSpace* attributes

Next, we list all vertex and edge attributes that can be passed to *RGraphSpace* methods.

## Vertex attributes

```{r Node attributes, eval=TRUE, message=FALSE}
# Node color (Hexadecimal or color name)
V(gtoy1)$nodeColor <- c("red", "#00ad39", "grey80", "lightblue", "cyan")

# Node transparency (in [0,1])
V(gtoy1)$nodeAlpha <- 1

# Node size (numeric in [0, 100], as '%' of the plot space)
V(gtoy1)$nodeSize <- c(8, 5, 5, 10, 5)

# Node shape (integer code between 0 and 25; see 'help(points)')
V(gtoy1)$nodeShape <- c(21, 22, 23, 24, 25)

# Node line width (as in 'lwd' standard graphics; see 'help(gpar)')
V(gtoy1)$nodeLineWidth <- 1

# Node line color (Hexadecimal or color name)
V(gtoy1)$nodeLineColor <- "grey20"

# Node labels ('NA' will omit the label)
V(gtoy1)$nodeLabel <- c("V1", "V2", "V3", "V4", NA)

# Node label size (in mm)
V(gtoy1)$nodeLabelSize <- 3

# Node label color (Hexadecimal or color name)
V(gtoy1)$nodeLabelColor <- "black"
```

## Edge attributes

Given a list of edges, *RGraphSpace* represents only one edge for each pair of connected vertices. If there are multiple edges connecting the same node pair, it will display the attributes of the first occurrence in the data.

```{r Edge attributes - 1, eval=TRUE, message=FALSE}
# Edge color (Hexadecimal or color name)
E(gtoy1)$edgeColor <- c("red","green","blue","black")

# Edge transparency (in [0,1])
E(gtoy1)$edgeAlpha <- 1

# Edge line width (as in 'lwd' standard graphics; see 'help(gpar)')
E(gtoy1)$edgeLineWidth <- 0.8

# Edge line type (as in 'lty' standard graphics; see 'help(gpar)')
E(gtoy1)$edgeLineType <- c("solid", "11", "dashed", "2124")
```

Note: `edgeLineColor` is deprecated as of version 1.4.3 and replaced by `edgeColor`.

## Arrowhead attributes

**Arrowhead in directed graphs**: By default, an arrow will be drawn for each edge according to its left-to-right orientation in the edge list (*e.g.* `A -> B`). If there are mutual connections, the package will recode the mutual edges to represent a bidirectional flow.

```{r Edge attributes - 2, eval=TRUE, message=FALSE}
# Arrowhead types in directed graphs
## Integer or character code:
## 0 = "---", 1 = "-->", -1 = "--|"
E(gtoy1)$arrowType <- 1
``` 

**Arrowhead in undirected graphs**: By default, no arrow will be drawn for undirected graphs. However, arrowheads may be assigned according to the coding below.

```{r Edge attributes - 3, eval=TRUE, message=FALSE}
# Arrowhead types in undirected graphs
## Integer or character code:
##  0 = "---"
##  1 = "-->",  2 = "<--",  3 = "<->",  4 = "|->"
## -1 = "--|", -2 = "|--", -3 = "|-|", -4 = "<-|"
gtoy1_undir <- igraph::as_undirected(gtoy1, edge.attr.comb = "first")
E(gtoy1_undir)$arrowType <- 1
# Note: in undirected graphs, this attribute overrides
# the edge's orientation in the edge list and adds arrowheads
# to edges that would otherwise be drawn without any
```

... and plot the fully attributed `gtoy1` object.
 
```{r A shortcut for RGraphSpace, eval=TRUE, message=FALSE}
# Plot the fully attributed 'gtoy1'
plotGraphSpace(gtoy1, node.labels = TRUE)
```

```{r, eval=FALSE, message=FALSE, echo=FALSE, include=FALSE, purl=FALSE}
# p <- plotGraphSpace(gtoy1, node.labels = TRUE)
# ggsave(filename = "./vignettes/cards/building_gspace.png", height=4, width=4,
#   units="in", device="png", dpi=200, plot = p)
```

# Passing graphs to *geoms*

Alternatively, an `igraph` can be converted to a `GraphSpace` object and passed directly to *ggplot2* geoms. This gives full access to the *ggplot2* layer system for combining graph elements with other plot types.

```{r Using geoms - 1, eval=TRUE, message=TRUE}
# Load the toy graph used in the previous example
data("gtoy1", package = "RGraphSpace")

# Create a GraphSpace object
gs <- GraphSpace(gtoy1)

# Normalize the coordinates
gs <- normalizeGraphSpace(gs)

gs
```

`normalizeGraphSpace()` maps all vertex coordinates to a `[0, 1]` unit interval and computes the per-node clipping offsets that allow `geom_edgespace()` to terminate edges precisely at node boundaries. This step is handled automatically when you use `plotGraphSpace()`, but must be called explicitly when building a plot layer by layer.

```{r Using geoms - 2, eval=TRUE, message=FALSE}
# Build a layered ggplot2 graph
# geom_edgespace() draws edges; geom_nodespace() draws nodes
# aes(label = nodeLabel) maps the 'nodeLabel' vertex attribute to node labels
ggplot(gs) + 
  geom_edgespace() + 
  geom_nodespace(aes(label = nodeLabel), label_size = 3) + 
  theme_gspace_coords(is_norm = TRUE)
```

For detailed integration with the *ggplot2* ecosystem and other spatial packages, see [*customizing aesthetics*](customizing-aesthetics.html) and [*interoperability with ggraph & sf*](interoperability.html) vignettes.


# Choosing an entry point

*RGraphSpace* provides three levels of access, each suited to a different workflow.

**Level 1 — Direct plot from an `igraph`**: The simplest call requires no intermediate objects. Use this for quick inspection or when no *ggplot2* customization is needed.

```{r Entry point 1, eval=FALSE}
# The simplest call
plotGraphSpace(gtoy1, node.labels = TRUE)
```

**Level 2 — Layered plot via `geom_graphspace()`**: Convert the `igraph` to a `GraphSpace` object first, then pass it to *ggplot2*. `geom_graphspace()` adds node and edge layers in a single call and gives full access to *ggplot2* themes, scales, and annotations.

```{r Entry point 2, eval=FALSE}
# Adds node and edge layers in a single call
gs <- GraphSpace(gtoy1)
ggplot(gs) +
  geom_graphspace(aes(label = nodeLabel))
```

**Level 3 — Independent node and edge layers**: Use `geom_nodespace()` and `geom_edgespace()` directly when node and edge layers require separate aesthetic mappings or independent scale control.

```{r Entry point 3, eval=FALSE}
# Set some variables
V(gtoy1)$node_var <- runif(vcount(gtoy1))
E(gtoy1)$edge_var <- runif(ecount(gtoy1))

# Independent node and edge layers
gs <- GraphSpace(gtoy1)
ggplot(gs) +
  geom_edgespace(aes(colour = edge_var)) +
  geom_nodespace(aes(fill = node_var))
```

The `layout` argument of `GraphSpace()` can be used at levels 2 and 3 to supply coordinates from any *igraph* layout algorithm when the graph has no pre-existing `x` and `y` vertex attributes:

```{r Entry point layout, eval=FALSE}
# Entry point layout
gs <- GraphSpace(gtoy1, layout = igraph::layout_with_fr(gtoy1))
```

# Session information
```{r label='Session information', eval=TRUE, echo=FALSE}
sessionInfo()
```

