---
title: "Curved Edges, Parallel Edges, and Self-Loops"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Curved Edges, Parallel Edges, and Self-Loops}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

```{r Check required version}
# Check required version
if (packageVersion("RGraphSpace") < "1.5.4"){
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}
```

## Overview

The `geom_edgespace()` function can render edges as curves through its `curve` argument. Curved edges are especially useful when a pair of nodes is connected by more than one edge, or when an edge starts and ends at the same node (a self-loop).

## Setting up the example data

We build a small toy graph containing ordinary edges, parallel edges, and self-loops. A self-loop is an edge whose two endpoints are the same vertex, whereas parallel edges are edges that connect the same pair of vertices. The graph is wrapped into `GraphSpace` objects with `simplify = FALSE`, since parallel edges and self-loops are otherwise removed by default when a graph is validated.

```{r Load packages, eval=TRUE, message=FALSE}
# Load packages
library("RGraphSpace")
library("igraph")
library("ggplot2")
```

```{r Setting up the example data, eval=TRUE, message=TRUE}
# Toy graph topology: hub + ring
n <- 10; hub <- 5
g <- make_ring(n, directed = TRUE) %u% 
  make_star(n, center = hub, mode = "out")
  
# Add self-loops
g <- add_edges(g, rep(c(2,4,6,8,10,10), each = 2))

# Add parallel edges
g <- add_edges(g, c(6,5, 6,5, 5,4))

# Set node and edge colors via igraph attributes directly, rather than
# through a geom, as these values will be fixed (not data-driven)
V(g)$nodeColor <- "steelblue1"
E(g)$edgeColor <- "purple"
E(g)$label <- seq_len(ecount(g))
  
# Build a GraphSpace with a circular layout; simplify = FALSE keeps the
# parallel edges and loops intact
gs_toy <- GraphSpace(g, layout = layout_in_circle(g), simplify = FALSE)

# Rescale coordinates to a standard range 
# (optional, but recommended for consistent plotting)
gs_toy <- normalizeGraphSpace(gs_toy)

gs_toy
```

## Curved edges

By default, simple edges are drawn as straight lines.

```{r Straight edges, eval=TRUE, message=TRUE}
ggplot(gs_toy) +
  geom_edgespace(aes(label = label), label_size = 2) +
  geom_nodespace(aes(label = nodeLabel), label_size = 3) +
  theme_gspace_coords(is_norm = TRUE)
```

Increasing `curve` bows simple edges away from a straight line. Arrowheads, labels, and node-clipping offsets automatically follow the local `curve` direction, so they stay correctly oriented as the edge bends.

```{r Gentle curve, eval=TRUE, message=TRUE}
ggplot(gs_toy) +
  geom_edgespace(aes(label = label), label_size = 2, curve = 0.2) +
  geom_nodespace(aes(label = nodeLabel), label_size = 3) +
  theme_gspace_coords(is_norm = TRUE)
```

The sign of `curve` sets which side the simple edge bows toward.

```{r Curve sign flips the side, eval=TRUE, message=TRUE}
ggplot(gs_toy) +
  geom_edgespace(aes(label = label), label_size = 2, curve = -0.2) +
  geom_nodespace(aes(label = nodeLabel), label_size = 3) +
  theme_gspace_coords(is_norm = TRUE)
```

## Parallel edges and self-loops

`curve` applies uniformly to simple edges in a layer, which is not enough on its own to keep parallel edges apart, or to make a self-loop visible (a loop has no length for `curve` to be a fraction of). *RGraphSpace* handles both automatically: edges that share the same pair of nodes fan out, and self-loops are drawn with their own geometry -- all without any extra arguments.

The amount of fan-out is controlled by a separate argument, `parallel_spread` (default `1`), independently of `curve`. Lower values keep parallel edges and loops closer together; higher values spread them further apart.

```{r Curve and parallel_spread together, eval=TRUE, message=TRUE}
ggplot(gs_toy) +
  geom_edgespace(aes(label = label), label_size = 2,
    curve = 0.3, parallel_spread = 2) +
  geom_nodespace(aes(label = nodeLabel), label_size = 3) +
  theme_gspace_coords(is_norm = TRUE)
```

## Orienting self-loops

Self-loops also accept a `loop_direction` argument, which controls where a loop sits relative to its node, independently of `parallel_spread`'s control over its size. The default, `"adaptive"`, points each loop away from the graph's centroid, so loops naturally expand toward less crowded regions of the layout. A single numeric value fixes every loop at the same angle (in degrees), regardless of node position.

```{r Adaptive loop direction, eval=TRUE, message=TRUE}
ggplot(gs_toy) +
  geom_edgespace(aes(label = label), label_size = 2, 
    curve = 0.3, parallel_spread = 2, loop_direction = 90) +
  geom_nodespace(aes(label = nodeLabel), label_size = 2) +
  theme_gspace_coords(is_norm = TRUE)
```


```{r, eval=FALSE, message=FALSE, echo=FALSE, include=FALSE, purl=FALSE}
# p <- ggplot(gs_toy) +
#   geom_edgespace(aes(label = label), label_size = 2,
#     curve = 0.3, parallel_spread = 2) +
#   geom_nodespace(aes(label = nodeLabel), label_size = 3) +
#   theme_gspace_coords(is_norm = TRUE)
# ggsave(filename = "./vignettes/articles/cards/curved_edges.png", height=4, width=4,
#   units="in", device="png", dpi=200, plot = p)
```

## Session information

```{r Session information}
sessionInfo()
```
