---
title: "Customizing Aesthetics"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width="90%"
  )
```

<br/>
**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

```{r Check required version, eval=TRUE, message=FALSE}
# Check required version
if (packageVersion("RGraphSpace") < "1.5.2"){
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}
```

# Overview

This section illustrates how *RGraphSpace* integrates with *ggplot2* using its building blocks [@Wickham2016]. Graph attributes stored in the `GraphSpace` object can be handled in two ways:

* **Identity mapping:** Attributes such as `nodeColor`, `nodeSize`, and `nodeShape` are treated as literal values and displayed exactly as specified, without scaling or transformation.

* **Dynamic aesthetic mapping:** Attributes are mapped to *ggplot2* aesthetics such as `colour`, `size`, and `shape`, and rendered through standard scales, which automatically generate synchronized legends.

*RGraphSpace* implements three specialized `geoms` for handling graph data within a *ggplot2* workflow. These `geoms` synchronize node and edge layers, which is essential when network elements must remain accurately aligned with a reference frame.

1. **`geom_nodespace()`**: Renders network nodes. Extends `GeomPoint` aesthetic mappings and exposes node state information to the edge layer.

2. **`geom_edgespace()`**: Renders the relationships between nodes. Extends `GeomSegment` aesthetic mappings; unlike standard segments, it is node-aware and dynamically adjusts start and end points based on node position and size.

3. **`geom_graphspace()`**: A convenience wrapper that calls `geom_nodespace()` and `geom_edgespace()` in sequence. Use this for the common case; use the individual `geoms` directly when independent control of node and edge layers is needed.

# Setting basic input data

In the following example, we create a small modular graph containing variables of different types to demonstrate how *RGraphSpace* `geoms` handle different mapping requirements.

```{r Load packages, eval=TRUE, message=FALSE}
#--- Load packages
library("RGraphSpace")
library("igraph")
library("ggplot2")
```

```{r Load a toy graph, eval=TRUE, message=TRUE}
# Make a toy modular graph
set.seed(42)
gtoy2 <- sample_islands(
  islands.n = 3,       # number of modules
  islands.size = 30,   # nodes per module
  islands.pin = 0.25,  # probability of edges within modules
  n.inter = 2)         # edges between modules

# Assign module membership to nodes
V(gtoy2)$module <- rep(1:3, each = 30)

# Assign colors to nodes
V(gtoy2)$nodeFillColor <- rainbow(3)[V(gtoy2)$module]

# Assign a categorical variable to nodes
V(gtoy2)$node_group <- c("A", "B", "C")[V(gtoy2)$module]

# Assign numeric variables to nodes and edges
V(gtoy2)$node_var <- runif(vcount(gtoy2))
E(gtoy2)$edge_var <- runif(ecount(gtoy2))

# Create a GraphSpace from the toy igraph
gs <- GraphSpace(gtoy2)

gs
```

# Plotting identity values

In this example, `nodeColor` already contains the final colour values stored in the `GraphSpace` object. The colours will be displayed as-is by the `geom_graphspace()` function. This approach is useful when nodes have been pre-processed with specific attributes and you want the visual output without further mapping.

```{r, eval=TRUE, message=FALSE, include = FALSE}
edge_var <- node_var <- nodeFillColor <- node_group <- NULL
```

```{r Plot identity values, eval=TRUE, message=FALSE}
ggplot(gs) + 
  geom_graphspace() +
  theme(aspect.ratio = 1)
```

The trade-off of this approach is that all attributes reflect the original data directly, but no legend is accessible. This is because identity scales bypass the legend-building process of *ggplot2*. If a legend is required to explain the meaning of these colours, the attribute should be mapped via aesthetics (e.g., `aes(fill = attribute)`) and modified by a `scale_*()` function.

# Mapping categorical variables

In this example, the node categorical variable `node_group` is mapped to the `fill` aesthetic and we use the individual `geoms` directly for independent control of node and edge layers.

```{r Map aesthetics to categorical variables, eval=TRUE, message=FALSE}
ggplot(gs) + 
  geom_edgespace() +
  geom_nodespace(aes(fill = node_group), colour = "grey") +
  scale_fill_viridis_d(option = "viridis") +
  theme_gspace_coords()
```

```{r, eval=FALSE, message=FALSE, echo=FALSE, include=FALSE, purl=FALSE}
# p <- ggplot(gs) + geom_edgespace() +
#   geom_nodespace(aes(fill = node_group), colour = "grey") +
#   scale_fill_viridis_d(option = "viridis") +
#   theme_gspace_coords()
# ggsave(filename = "./vignettes/cards/using_geoms.png", height=4, width=4,
#   units="in", device="png", dpi=200, plot = p)
```

# Mapping numeric variables

In this example, node and edge numeric variables are mapped to `fill` and `colour` aesthetics, respectively.

```{r Map aesthetics to numeric variables, eval=TRUE, message=FALSE}
# Map aesthetics to numeric variables
ggplot(gs) + 
  geom_edgespace(aes(colour = edge_var)) +
  geom_nodespace(aes(fill = node_var), colour = "grey") +
  scale_colour_continuous(palette = c("cyan","blue")) +
  scale_fill_continuous(palette = c("white","purple")) +
  theme_gspace_coords()
```

# Using separate colour scales

If you need to map different variables to the same aesthetic (such as `colour`) with independent scales, the *ggnewscale* package offers an elegant solution to introduce additional scales within the same plot [@Campitelli2025]; for example:

```{r Map aesthetics to separate colour scales, eval=TRUE, message=FALSE}
if (!require("ggnewscale", quietly = TRUE)) {
  install.packages("ggnewscale")
}
library("ggnewscale")
ggplot(data = gs) + 
  geom_edgespace(aes(colour = edge_var)) +
  scale_colour_continuous(palette = c("cyan","blue")) +
  ggnewscale::new_scale_colour() +
  geom_nodespace(aes(colour = node_var), 
    stroke = 2, fill = NA) +
  scale_colour_continuous(palette = c("white","purple")) +
  theme_gspace_coords()
```

# Session information
```{r label='Session information', eval=TRUE, echo=FALSE}
sessionInfo()
```

# References

