---
title: "Using 'sf' geometries with RGraphSpace"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
  )
```

<br/>

**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

```{r Check required version, eval=TRUE, message=FALSE}
# Check required version
if (packageVersion("RGraphSpace") < "1.5.5"){
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}
```

# Overview

This tutorial demonstrates how *sf* geometries can be attached to graph nodes, then either fit to the nodes as markers or normalized into a shared coordinate space, using `GraphSpace` geometry accessors.

## Setting basic input data

We first construct a star-like network with nodes of different sizes. The graph is converted to a `GraphSpace` object and node coordinates are normalized.

```{r Load packages, eval=TRUE, message=FALSE}
#--- Load packages
if (!requireNamespace("sf", quietly = TRUE)) {
    install.packages("sf")
}
library("RGraphSpace")
library("igraph")
library("ggplot2")
library("sf")
library("patchwork")
```

```{r Geometry - 1, eval=TRUE, message=FALSE}
# Make a toy graph
gtoy_star <- make_star(20, mode="out")

# Make a 'GraphSpace'
gs_star <- GraphSpace(gtoy_star, layout = layout_as_star(gtoy_star))

# Set node color and size
gs_star$nodeFillColor <- adjustcolor("blue", 0.1)
gs_star$nodeSize <- seq(1, gs_vcount(gs_star) )

# Normalize node coordinates
gs_star <- normalizeGraphSpace(gs_star)
```

The default node markers already reflect each node's varying size.

```{r Geometry - 2, eval=TRUE, message=FALSE}
# Plot with ggplot2 and RGraphSpace geoms
ggplot(gs_star) + 
  geom_edgespace() +
  geom_nodespace(colour = "red") + 
  theme_gspace_coords(is_norm = TRUE)
```

</br>

Next, we build a set of decorative `sf` shapes, unrelated to the graph, with no inherent size or position of their own.

```{r Geometry - 3, eval=TRUE, message=FALSE}
# Make some decorative shapes
shapes20 <- c( sfshape_ngons(n = 10, sides = 3:7), 
  sfshape_stars(n = 10, points = 3:7) )

# Plot with ggplot2's geom_sf
ggplot(shapes20) + geom_sf() + 
  ggtitle("Decorative 'sf' geometries")
```

</br>

# Fit geometries to nodes

We can attach the geometries to the graph through the `gs_geometry()` accessor, and then use `fitGeometry()` to position and size each geometry to match its corresponding node.

```{r Geometry - 4, eval=TRUE, message=FALSE}
# Attach the decorative shapes to the graph nodes
gs_geometry(gs_star, "geometry") <- shapes20

# Fit each shape to its node's size and position
gs_star <- fitGeometry(gs_star)
```

The geometries are now aligned with their nodes; positions follow the node coordinates, and sizes match the node sizes.

```{r Geometry - 5, eval=TRUE, message=FALSE}
# Plot nodes and fitted geometries
ggplot(gs_star) + 
  geom_edgespace() +
  geom_nodespace(colour = "red") +
  geom_sf(aes(geometry = geometry), fill = "blue") +
  theme_gspace_coords(is_norm = TRUE) +
  ggtitle("Decorative 'sf' geometries fit to nodes")
```

```{r, eval=FALSE, message=FALSE, echo=FALSE, include=FALSE, purl=FALSE}
# p <- ggplot(gs_star) +
#   geom_edgespace() +
#   geom_nodespace(colour = "red") +
#   geom_sf(aes(geometry = geometry), fill = "blue") +
#   theme_gspace_coords(is_norm = TRUE)
# ggsave(filename = "./vignettes/articles/cards/geometry.png", height=4, width=4,
#   units="in", device="png", dpi=200, plot = p)
```

# Geometry normalization

In the previous section, `fitGeometry()` positioned and sized decorative shapes to match their nodes. That works because the shapes had no coordinate space of their own. When geometries *do* carry their own coordinates, `fitGeometry()` alone cannot align them.

Consider a scenario in which a tissue sample is both imaged and sampled: the samples give the node coordinates, the image structures give the geometries. Both map the same tissue, so they share relative positions, but having been mapped independently, they no longer share a common scale.

The following example illustrates this scenario: nodes and geometries correspond in position but differ in scale. Node sizes are reset to a flat value, so we plot them over the geometries, where they stay visible within each shape.

```{r Geometry - 6, eval=TRUE, message=FALSE, out.width="100%"}
gs_star2 <- gs_star

# Reset nodeSize, so node-shape sizes no longer relate
gs_star2$nodeSize <- 2

p1 <- ggplot(gs_star2) + 
  geom_sf(aes(geometry = geometry), fill = "blue") +
  geom_edgespace() +
  geom_nodespace(fill = "cyan") +
  theme_gspace_coords(is_norm = TRUE) +
  ggtitle("Nodes and geometries\nwith unrelated sizes")

# Set a new scale factor to node coordinates, 
# so node-shape positions also no longer relate.
# Note: this denormalizes the coordinates.
gs_scale_factor(gs_star2) <- 0.2

p2 <- ggplot(gs_star2) + 
  geom_sf(aes(geometry = geometry), fill = "blue") +
  geom_edgespace() +
  geom_nodespace(fill = "cyan") +
  theme_gspace_coords(is_norm = FALSE) +
  ggtitle("Nodes and geometries\nin unrelated spaces")

p1 + p2
```

Next, we try `fitGeometry()` with `use_node_size = FALSE`, repositioning each shape onto its node without resizing. Since we reset `nodeSize` to a flat value, sizing to it would be meaningless, so only position can be recovered.

```{r Geometry - 7, eval=TRUE, message=FALSE}
# Fit shapes to node positions, not sizes
gs_star2_fit <- fitGeometry(gs_star2, use_node_size = FALSE)

ggplot(gs_star2_fit) + 
  geom_sf(aes(geometry = geometry), fill = "blue") +
  geom_edgespace() +
  geom_nodespace(fill = "cyan") +
  theme_gspace_coords(is_norm = FALSE) +
  ggtitle("Nodes and geometries aligned in position only")
```

Next, we use `normalizeGraphSpace(..., norm.geometry = TRUE)`, which normalizes the nodes and then calls `normalizeGeometry()` on the geometry. Unlike `fitGeometry()`, this recovers scale from the geometry itself: it fits a linear relationship between the geometry's centroids and the node coordinates, bringing both into one common space.

```{r Geometry - 8, eval=TRUE, message=FALSE}
# Normalize node coordinates and geometries
gs_star2_norm <- normalizeGraphSpace(gs_star2, norm.geometry = TRUE)

ggplot(gs_star2_norm) + 
  geom_sf(aes(geometry = geometry), fill = "blue") +
  geom_edgespace() +
  geom_nodespace(fill = "cyan") +
  theme_gspace_coords(is_norm = TRUE) +
  ggtitle("Nodes and geometries co-normalized")
```


# Session information
```{r label='Session information', eval=TRUE, echo=FALSE}
sessionInfo()
```

