---
title: "Using 'sf' geometries with RGraphSpace"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
  )
```

<br/>

**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

```{r Check required version, eval=TRUE, message=FALSE}
# Check required version
if (packageVersion("RGraphSpace") < "1.5.2"){
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}
```

# Overview

# Setting basic input data

Below, we construct a star-like network with varying node sizes to show how the geometries stay synchronized across a wide range values.

```{r Load packages, eval=TRUE, message=FALSE}
#--- Load packages
library("RGraphSpace")
library("igraph")
library("ggplot2")
library("sf")
```

```{r Geometry - 1, eval=TRUE, message=FALSE}
# Make a toy graph
gtoy_star <- make_star(20, mode="out")

# Make a 'GraphSpace'
gs_star <- GraphSpace(gtoy_star, layout = layout_as_star(gtoy_star))

# Set node color and size
gs_star$nodeFillColor <- adjustcolor("blue", 0.1)
gs_star$nodeSize <- seq(1, gs_vcount(gs_star) )

# Normalize node coordinates
gs_star <- normalizeGraphSpace(gs_star)
```

The default node markers already reflect each node's varying size:

```{r Geometry - 2, eval=TRUE, message=FALSE}
# Plot with ggplot2 and RGraphSpace geoms
ggplot(gs_star) + 
  geom_edgespace() +
  geom_nodespace(colour = "red") + 
  theme_gspace_coords(is_norm = TRUE)
```

</br>

Next, build a set of decorative `sf` shapes, unrelated to the graph, with no inherent size or position of their own:

```{r Geometry - 3, eval=TRUE, message=FALSE}
# Make some decorative shapes
shapes20 <- c( sfshape_ngons(n = 10, sides = 3:7), 
  sfshape_stars(n = 10, points = 3:7) )

# Plot with ggplot2's geom_sf
ggplot(shapes20) + geom_sf() + 
  ggtitle("Decorative 'sf' geometries")
```

</br>

Attach the shapes to the nodes, then let `fitGeometry()` position and size each one to match its node exactly:

```{r Geometry - 4, eval=TRUE, message=FALSE}
# Attach the decorative shapes to the graph nodes
gs_geometry(gs_star, "geometry") <- shapes20

# Fit each shape to its node's size and position
gs_star <- fitGeometry(gs_star)
```

The geometries now track each node's size and position precisely, plotted here alongside the original markers for comparison:

```{r Geometry - 5, eval=TRUE, message=FALSE}
# Plot nodes and fitted geometries
ggplot(gs_star) + 
  geom_edgespace() +
  geom_nodespace(colour = "red") +
  geom_sf(aes(geometry = geometry), fill = "blue") +
  theme_gspace_coords(is_norm = TRUE) +
  ggtitle("Decorative 'sf' geometries fit to nodes")
```

```{r, eval=FALSE, message=FALSE, echo=FALSE, include=FALSE, purl=FALSE}
# p <- ggplot(gs_star) +
#   geom_edgespace() +
#   geom_nodespace(colour = "red") +
#   geom_sf(aes(geometry = geometry), fill = "blue") +
#   theme_gspace_coords(is_norm = TRUE)
# ggsave(filename = "./vignettes/articles/cards/geometry.png", height=4, width=4,
#   units="in", device="png", dpi=200, plot = p)
```

# Session information
```{r label='Session information', eval=TRUE, echo=FALSE}
sessionInfo()
```

