---
title: "Getting Started with RGraphSpace"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output: 
  html_document:
    toc: false
---

```{css, echo=FALSE}
.card-title {
  text-align: center !important;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bs4cards)
```

<br/>
**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

# Introductory vignettes

These tutorials introduce *RGraphSpace* using simple toy examples: [*building a graph space*](building-graphspace.html) walks through basic graph conventions; [*customizing aesthetics*](customizing-aesthetics.html) demonstrates how to set up `geoms` to handle graph data types; and [*fine-tuning scales*](scales-and-offsets.html) describes the trade-offs involved in synchronizing node and edge layers.

```{r, include=FALSE}
title <- c("Building GraphSpace", "Customizing Aesthetics", 
  "Fine-tuning Scales")
image <- c("cards/building_gspace.png", "cards/using_geoms.png", "cards/scales_and_offsets.png")
link <- c("building-graphspace.html","customizing-aesthetics.html",
  "scales-and-offsets.html")
ps_cards1 <- data.frame(title = title, image = image, link = link)
```


```{r, echo=FALSE}
ps_cards1 %>% 
  cards(
      title = title,
      image = image,
      link = link,
      layout = "label-below", 
      width = 3,
      border_radius = 5
  )
```

---

<div class="mt-5 mb-4"></div>

# General applications

These tutorials showcase *RGraphSpace* applied to real-world scenarios: [*mapping graphs to images*](mapping-images.html) illustrates the use of images as spatial references; [*interoperability with ggraph & sf*](interoperability.html) details integration strategies; and [*interactive visualization*](interactive.html) demonstrates manual graph layout refinement.

```{r, include=FALSE}
title <- c("Graphs to Images", "Interoperability", "Interactive Visualization")
image <- c("cards/mapping_images.png", "cards/interoperability.png", "cards/interactive.png")
link <- c("mapping-images.html","interoperability.html","interactive.html")
ps_cards2 <- data.frame(title = title, image = image, link = link)
```


```{r, echo=FALSE}
ps_cards2 %>% 
  cards(
      title = title,
      image = image,
      link = link,
      layout = "label-below", 
      width = 3,
      border_radius = 5
  )
```

<div class="mt-5 mb-4"></div>

# Advanced workflows


```{r, include=FALSE}
title <- c("High-Dimensional Data", "Spatial Data")
image <- c("cards/high_dim.png", "cards/spatial_data.png")
link <- c("high-dimensional.html","spatial-data.html")
ps_cards3 <- data.frame(title = title, image = image, link = link)
```

```{r, echo=FALSE}
ps_cards3 %>% 
  cards(
      title = title,
      image = image,
      link = link,
      layout = "label-below", 
      width = 3,
      border_radius = 5
  )
```

<div class="mt-5 mb-4"></div>

# *PathwaySpace* workflows

These tutorials illustrate how *RGraphSpace* integrates with *PathwaySpace* to project and visualize network signals in a high-dimensional context.

* <a href="https://sysbiolab.github.io/PathwaySpace/" target="_blank">https://sysbiolab.github.io/PathwaySpace/</a>

