---
title: "GraphSpace accessors and transformations"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width="90%"
  )
```

<br/>
**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

```{r Check required version, eval=TRUE, message=FALSE}
# Check required version
if (packageVersion("RGraphSpace") < "1.5.4"){
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}
```

# Overview

*RGraphSpace* is primarily designed for graph rendering and expects graphs to be prepared before they enter the *RGraphSpace* workflow. More extensive graph transformations are therefore generally best performed upstream, while preparing the data used as input to *RGraphSpace*.

Nevertheless, downstream analyses may require accessing and modifying a `GraphSpace` object to explore alternative graph configurations or highlight specific subsets of the data.

In the following, we demonstrate the `GraphSpace` accessors for graph manipulation. All `GraphSpace` accessors use the `gs_*` prefix to reduce the chance of masking functions from other packages, particularly graph-analysis packages that are commonly used side-by-side.

## Setting basic input data

```{r Load packages for quick start, eval=TRUE, message=FALSE}
#--- Load required packages
library("RGraphSpace")
library("igraph")
library("ggplot2")
```

```{r , eval=TRUE, message=FALSE, results='hide'}
# Load a demo igraph
data('gtoy1', package = 'RGraphSpace')

# Create a new GraphSpace object
gs <- GraphSpace(gtoy1)
```

# Manipulating attributes

The `gs_vertex_attr()` and `gs_edge_attr()` functions provide a single interface to access, add, modify, and delete node and edges attributes, respectively. This interface wraps the corresponding *igraph* methods, applying them to the internal graph representation and propagating the result to all `GraphSpace` components.

## Node attributes

```{r , eval=TRUE, message=FALSE}
# Access all vertex attributes
# gs_vertex_attr(gs)

# Access a specific vertex attribute
gs_vertex_attr(gs, "nodeLabel")

# Modify a single value within a vertex attribute
gs_vertex_attr(gs, "nodeSize")["n1"] <- 10

# Replace an entire vertex attribute
gs_vertex_attr(gs, "nodeSize") <- 10

# Add a new vertex attribute
gs_vertex_attr(gs, "new_node_var") <- rnorm(gs_vcount(gs))

# Delete a vertex attribute by assigning NULL
gs_vertex_attr(gs, "new_node_var") <- NULL

gs
```

## Edge attributes

```{r , eval=TRUE, message=FALSE}
# Access a specific edge attribute
gs_edge_attr(gs, "edgeColor")

# Replace an entire edge attribute
gs_edge_attr(gs, "edgeLineWidth") <- 1

# Add a new edge attribute
gs_edge_attr(gs, "new_edge_var") <- rnorm(gs_ecount(gs))

# Delete an edge attribute by assigning NULL
gs_edge_attr(gs, "new_edge_var") <- NULL

gs
```

# Adding nodes

The `gs_add_nodes()` function adds one or more nodes to a `GraphSpace` object. Attributes present on existing nodes but absent in the input `value` are filled from package defaults. Standard node attributes (such as `nodeSize` and `nodeColor`) are kept consistent across old and new nodes; the` @graph`, `@nodes`, and `@fdata` slots are updated consistently. Because new nodes introduce coordinates into the existing layout, the normalized state is invalidated and `normalizeGraphSpace()` must be re-run afterwards.

```{r , eval=TRUE, message=FALSE}
# Functional form (pipe-friendly) returns a modified copy
gs <- gs |> gs_add_nodes(data.frame(name = "n6", x = 0.5, y = 0.5))

# Assignment form modifies gs in place
gs_add_nodes(gs) <- data.frame(name = "n7", x = 0.5, y = 0.5)

# Add multiple nodes with visual attributes
gs <- gs_add_nodes(gs, data.frame(
  name = c("n8", "n9"),
  x = c(0.5, 0.8),
  y = c(0.5, 0.2),
  nodeSize = c(8, 5),
  nodeColor = c("steelblue", "tomato")) )
```

# Adding edges

The `gs_add_edges()` function adds one or more edges to a `GraphSpace` object. Both endpoints of every new edge must already exist in the node set. The `@graph` and `@edges` slots, along with all derived edge quantities, are updated consistently; the node set and the normalized coordinate state are not affected.

```{r , eval=TRUE, message=FALSE}
# Functional form (pipe-friendly) returns a modified copy
gs <- gs |> gs_add_edges(data.frame(from = "n2", to = "n3"))

# Assignment form modifies gs in place
gs_add_edges(gs) <- data.frame(from = "n3", to = "n4")

# Add multiple edges with a numeric attribute
gs <- gs_add_edges(gs, data.frame(
  from   = c("n4", "n5"),
  to     = c("n5", "n6"),
  weight = c(0.8, 0.4)) )
```

For objects built with `simplify = TRUE` (the default), loop edges (`from == to`), parallel edges, and duplicate rows within `value` are dropped with a warning. To allow loops or parallel edges, rebuild the object with `GraphSpace(g, simplify = FALSE)`.

# Subsetting nodes

The `gs_subset_nodes()` function retains a subset of nodes and automatically removes any edge whose endpoint is no longer present. The result is propagated to all `GraphSpace` components.

```{r , eval=TRUE, message=FALSE}
# Subset by node name (character vector)
gs2 <- gs_subset_nodes(gs, c("n1", "n2", "n3"))

# Subset by integer position
gs2 <- gs_subset_nodes(gs, 1:5)

# Subset by predicate (data masking against @nodes columns)
gs2 <- gs_subset_nodes(gs, nodeSize > 5)

# Subset by pre-evaluated logical vector
keep <- gs$nodeSize > 5
gs2  <- gs_subset_nodes(gs, keep)

# Combining with pipes
gs2 <- gs |>
  gs_subset_nodes(nodeSize > 5) |>
  gs_subset_edges(weight > 0.3)
```

# Subsetting edges

The `gs_subset_edges()` retains a subset of edges without modifying the node set, and the result is propagated to all `GraphSpace` components.

```{r , eval=TRUE, message=FALSE}
# Subset by predicate on an edge attribute
gs3 <- gs_subset_edges(gs, weight > 0.5)

# Subset by endpoint names: 'name1' and 'name2' are columns in
# @edges and can be used directly inside any predicate expression
gs3 <- gs_subset_edges(gs, name1 == "n1" & name2 == "n2")

# Combining endpoint and attribute conditions
gs3 <- gs_subset_edges(gs, name1 == "n1" & weight > 0.5)

# By integer position
gs3 <- gs_subset_edges(gs, 1:3)

# By logical vector
gs3 <- gs_subset_edges(gs, gs_edges(gs)$weight > 0.5)
```

# Subscript operators

The `[` operator subsets a `GraphSpace` object along two independent dimensions: nodes (`i`) and edges (`j`). This differs from the usual data-frame convention, where `[i, j]` indexes rows and columns of a single table. Here, neither index subsets columns; both select graph entities directly. Omitting an index retains all elements along that dimension.

**Synchronization rules:**

- `x[i, ]` **Node-induced subgraph.** Nodes are selected by `i`, then edges are automatically pruned to those whose endpoints both survived. Normalized coordinates are preserved.
- `x[, j]` **Edge selection.** Edges are selected by `j`; the node set is untouched and no node pruning occurs.
- `x[i, j]` **Combined selection.** Node filtering is applied first, then `j` is evaluated against the **original**, unfiltered edge table; an edge survives only if it satisfies `j` **and** both its endpoints survived node filtering.

```{r , eval=TRUE, message=FALSE}
# Node-induced subgraph: keep named nodes, prune dangling edges
gs[c("n1", "n2", "n3"), ]

# Node-induced subgraph by integer position
gs[1:4, ]

# Node-induced subgraph by pre-evaluated logical mask
gs[gs$nodeSize > 5, ]

# Edge selection only: keep all nodes
gs[, gs_edges(gs)$weight > 0.5]

# Edge selection by endpoint: predicates must be pre-evaluated
gs[, gs_edges(gs)$name1 == "n1"]

# Alternatively, wrap the predicate in quote()
gs[, quote(name1 == "n1" & name2 == "n2")]

# Combined: node filter first, then edge intersection
gs[c("n1", "n2", "n3"), gs_edges(gs)$weight > 0.5]
```

The `[[` operator, by contrast, is a simple accessor: `x[["nodes"]]`, `x[["edges"]]`, `x[["graph"]]`, and `x[["fdata"]]` return the corresponding component unmodified, equivalent to `getGraphSpace(x, "nodes")` and so on. No subsetting or synchronization logic applies.

```{r , eval=TRUE, message=FALSE, results='hide'}
gs[["nodes"]]   # same as getGraphSpace(gs, "nodes")
gs[["edges"]]   # same as getGraphSpace(gs, "edges")
gs[["graph"]]   # same as getGraphSpace(gs, "graph")
gs[["fdata"]]   # same as getGraphSpace(gs, "fdata")
```


# The `$` accessor

The `$` operator provides direct, attribute-style access to a single **vertex attribute** of a `GraphSpace` object, similar to how `$` works on a data frame or list. A companion `.DollarNames` method enables tab-completion after `gs$` in interactive sessions (e.g. RStudio), listing the vertex attributes currently defined on the graph.

```{r , eval=TRUE, message=FALSE}
# Return the vertex attribute `name`
gs$name

# Sets the vertex attribute `nodeShape`
gs$nodeShape <- 21
gs$nodeShape[1] <- 19

# Protected attributes cannot be modified this way
# gs$vertex <- 1  # error: 'vertex' is a read-only node attribute.
```

**Protected attributes**: those the package relies on internally (e.g. node `name` and `vertex`) are not editable through `$<-`; attempting to assign to one raises an error rather than silently modifying it. Reading via `gs$` is unaffected.


# General accessors

Here we show typical usage of the general accessors, used across the other tutorials.

```{r , eval=TRUE, message=FALSE, results='hide'}
# Vertex names
names(gs)

# Vertex attribute names
gs_names(gs)

# Get the node data frame
gs_nodes(gs)

# Get the edge data frame
gs_edges(gs)

# Get the underlying igraph object
gs_graph(gs)

# Number of vertices
gs_vcount(gs)

# Number of edges
gs_ecount(gs)

# Images may be provided as raster or numeric matrices;
# 'SpatRaster' objects are supported when the optional 
# 'terra' package is available
gs_image(gs) <- as_colorraster(volcano)

# Set a pixel budget for image operations
gs_image_maxpixels(gs) <- 4e+06

# Apply a scaling factor to node coordinates
gs_scale_factor(gs) <- 0.1

# Undo scaling
gs_scale_factor(gs) <- 1

# Normalize image and node coordinates to graph space
gs <- normalizeGraphSpace(gs, image.space = FALSE)

# Add a sparse Matrix aligned to nodes
library(Matrix)
mtx <- Matrix::Matrix(0, gs_vcount(gs), 2)
rownames(mtx) <- names(gs)
colnames(mtx) <- c("feature1","feature2")
gs_fdata(gs) <- mtx

# Feature names
gs_features(gs)

# Feature count
gs_nfeatures(gs)

# Add an 'sfc' geometry column (requires the optional 'sf' package)
if (requireNamespace("sf", quietly = TRUE)) {
  gs_geometry(gs) <- sfshape_ngons(n = gs_vcount(gs))
}
```


# Applying *igraph* functions

The `gs_compute()` function runs any igraph function on the graph carried by a `GraphSpace`, without needing a dedicated `gs_*` wrapper for each one. It extracts the underlying igraph via `as.igraph()`, applies `.f`, and returns the result unchanged. This is the read-only lane onto the broader igraph ecosystem: measures such as `degree()`, `betweenness()`, `coreness()`, and distances all work through this single entry point.

```{r , eval=TRUE, message=FALSE}
# Apply igraph functions
gs_compute(gs, igraph::degree)

gs_compute(gs, "betweenness", directed = FALSE)

# Fold a per-vertex result back as a node attribute
gs$degree <- gs_compute(gs, igraph::degree)
```

It is deliberately *not* a GraphSpace-modification path. If `.f` returns a graph (e.g. `simplify()`, `induced_subgraph()`), this cannot be reintegrated to the `gs` object, as graph modifications may invalidate the correspondence between the graph and other components of the object. A modified graph must instead go through the graph-modification checks of the `GraphSpace` constructor.

# Crop, rotate, flip, and transpose

These functions are special accessors, as they operate on a reference frame, either the graph or image space, not on the graph alone.

- `cropGraphSpace()` subsets a normalized `GraphSpace` to a specific region defined by the cropping boundaries. It recalculates node positions and background image boundaries to maintain spatial consistency after cropping, and drops nodes (and edges) that fall outside the window.
- `rotateGraphSpace()`, `flipGraphSpace()`, and `transposeGraphSpace()` are all exact coordinate/pixel permutations, with no resampling, no interpolation, and no risk of misaligning nodes against the background image. Applied to a normalized `GraphSpace`, they transform the normalized coordinates, leaving the raw coordinates intact so the original orientation can be restored without loss. Applied to a `GraphSpace` that has not been normalized, there are no normalized coordinates to act on, so the transformation is applied to the raw coordinates and is permanent. The `persist` argument controls this behavior: `persist = TRUE` writes the transformation into the raw coordinates regardless of normalization state.


```{r , eval=TRUE, message=FALSE}
# Create a GraphSpace
gs <- GraphSpace(make_full_graph(30))

# Normalize coordinates
gs <- normalizeGraphSpace(gs)

# Crop
gs_crop <- cropGraphSpace(gs, ymax = 0.5)

# Rotate
gs_rot90 <- rotateGraphSpace(gs)

# Flip
gs_flip <- flipGraphSpace(gs)

# Transpose
gs_t <- transposeGraphSpace(gs)

# Rotate
gs_rot90 <- rotateGraphSpace(gs, persist = TRUE)
```


# Session information
```{r label='Session information', eval=TRUE, echo=FALSE}
sessionInfo()
```

