---
title: "Mapping Graphs to Images"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
  )
```

<br/>
**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

```{r Check required version, eval=TRUE, message=FALSE}
# Check required version
if (packageVersion("RGraphSpace") < "1.5.0"){
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}
```

# Overview

Images can be used as spatial references for graphs. When a raster image is provided, the pixel coordinates define where nodes are positioned, supporting the construction of graphs from image features. However, spatial alignment requires consistent coordinate conventions between the graph and the image; a mismatch in axis orientation, for example a top-left versus bottom-left origin, leaves the nodes reflected relative to the image.

# Graph and image conventions

We begin with a simple toy example that highlights how graphs and images follow different conventions and need to be transformed into the same coordinate space for layered visualization. This toy illustrates a scenario in which the image registration has already been performed (see *image registration* section), with graph nodes representing image features.

```{r Coordinate conventions - 1, eval=TRUE, message=FALSE}
library("RGraphSpace")
library("igraph")
library("ggplot2")
library("patchwork")

# Load a toy GraphSpace object, which already
# includes an embedded image, with node 
# coordinates mapping to image indices.
data("gs_image_toy")

# Check node and image spatial boundaries
gs_image_toy
```

Node coordinates fall within the image's spatial boundaries: nodes range `x:[194, 493]`(cols) and `y:[237, 412]`(rows), while the image spans `x:[1, 680]`(cols) and `y:[1, 640]`(rows), consistent with a common frame of reference. The image is composed of several amorphous shapes and a bluish feature region to which the graph nodes are registered. A red node is included to break visual symmetry, and a corresponding reddish mark in the feature region allows its position to be tracked against the image.

```{r Coordinate conventions - 2, eval=TRUE, message=FALSE, fig.width=6, fig.height=3, out.width="100%"}
# Plot the un-normalized image
p1 <- ggplot() +
  annotation_gspace_image(gs_image_toy) +
  scale_x_continuous(name = "Image coordinates 1", limits = c(0, 1)) +
  scale_y_continuous(name = "Image coordinates 2", limits = c(0, 1)) +
  theme(aspect.ratio = 1)

# Plot the un-normalized graph
p2 <- ggplot(gs_image_toy) +
  geom_edgespace() + geom_nodespace() +
  scale_x_continuous(name = "Graph coordinates 1", limits = c(190, 500)) +
  scale_y_continuous(name = "Graph coordinates 2", limits = c(190, 500)) +
  theme(aspect.ratio = 1)

p1 + p2
```

When rendered as separate plots, the image spans `[0,1]`, while the graph spans `~[190, 500]`; these are different coordinate spaces entirely, so a simple merge would misplace them. A separate mismatch is orientation: the red node and its corresponding reddish mark in the feature region show that the image is rendered top-down, while the graph is rendered bottom-up.

Next, `normalizeGraphSpace()` maps node coordinates to image space and converts the source image into a render-ready canvas. By default, it flips node y-coordinates over the image center to match the image's orientation, then crops the image to the graph's extent (plus a margin), so that the resulting canvas is centered on the graph.

```{r Coordinate conventions - 3, eval=TRUE, message=TRUE}
gs_image_toy <- normalizeGraphSpace(gs_image_toy, mar = 0.1)

ggplot(gs_image_toy) +
  annotation_gspace_image(gs_image_toy) +
  geom_edgespace() + geom_nodespace() +
  theme_gspace_coords(is_norm = TRUE)
```

The default `mar = 0.1` leaves a margin of `0.1` on each side, so the graph occupies the central `0.8` of the canvas.

For reference, here is the result when **no** flip is applied to node coordinates:

```{r Coordinate conventions - 4, eval=TRUE, message=TRUE}
gs_image_toy <- normalizeGraphSpace(gs_image_toy, flip.y = FALSE)

ggplot(gs_image_toy) +
  annotation_gspace_image(gs_image_toy) +
  geom_edgespace() + geom_nodespace() +
  theme_gspace_coords(is_norm = TRUE)
```

**Note on image alignment**: Spatial misalignment may occur if the input image and node coordinates differ in axis orientation (e.g., top-left versus bottom-left origins). To accommodate these differences, `normalizeGraphSpace()` provides orientation controls through the `swap.xy`, `flip.x`, and `flip.y` arguments. If the nodes appear misaligned with the input image, try combinations of these parameters to correct the alignment. Alternatively, try `flip.v` and `flip.h` arguments to apply flipping directly to the background image (see the *orientation controls* subsection)


# Image registration

Image registration is a broad field concerned with estimating a spatial transformation that aligns a query to a reference dataset [@Lewis2021; @Balluff2022]. *RGraphSpace* addresses a constrained case where the spatial correspondence between graph and image is known by construction, so the alignment task reduces to correcting axis-orientation mismatches, fixing node positions to their target pixels, and normalizing graph and image to a common coordinate space. In this section we exemplify this constrained workflow by extracting pixel coordinates from an image matrix and building a graph with nodes placed at those positions.

## Setting basic input data

Next, we extract pixel coordinates at a specific intensity quantile from the `volcano` matrix and prepare an `igraph` for *RGraphSpace*.

```{r Mapping images - 1, eval=TRUE, message=FALSE}
library("RGraphSpace")
library("igraph")
library("ggplot2")

# Extract pixel coordinates for a specific intensity quantile.
coords <- which(volcano == quantile(volcano, 0.85), arr.ind = TRUE)

# Mark target pixels with '0'; it will appear as black in the background. 
# This creates a visual anchor to verify the alignment precision.
volcano2 <- volcano
volcano2[coords] <- 0

# Create an igraph object from the pixel coordinates; 
# note that at this stage, 'y' represents matrix row indices.
gtoy3 <- igraph::make_empty_graph(n = nrow(coords))
igraph::V(gtoy3)$y <- coords[,1]
igraph::V(gtoy3)$x <- coords[,2]

# Highlight the bottom-row vertex (max 'y' index) to demonstrate alignment; 
# since matrix indexing is top-down, this accounts for the default flip 
# between matrix and plot coordinate systems.
igraph::V(gtoy3)$nodeColor <- NA
bottom_row <- which.max(igraph::V(gtoy3)$y)
igraph::V(gtoy3)$nodeColor[bottom_row] <- adjustcolor("red", 0.4)
```

## Initialize, normalize, and plot

```{r Mapping images - 2, eval=TRUE, message=TRUE}
# Initialize a GraphSpace object
gs <- GraphSpace(gtoy3)

# Add a raster image
gs_image(gs) <- as_colorraster(volcano2)

# Map graph coordinates to the image space; by default,
# y-coordinates will be flipped (see comments below).
gs <- normalizeGraphSpace(gs)

gs
```

Observe that the nodes (open circles) are precisely aligned with the pixels (dark squares) to which they were mapped. In this plot, nodes remain proportional to the viewport regardless of the final output resolution. This behavior is especially critical when the network elements represent image features at specific positions and must stay anchored to those features, avoiding overlap with adjacent, unrelated areas.

```{r Mapping images - 3, eval=TRUE, message=FALSE}
# Render the graph with the raster as background
plotGraphSpace(gs, add.image = TRUE)
```

```{r Mapping images - 4, eval=TRUE, message=FALSE}
# Alternatively, passing to the underlying geoms
ggplot(gs) +
  annotation_gspace_image(gs) +
  geom_nodespace() +
  theme_gspace_coords(is_norm = TRUE)
```

```{r, eval=FALSE, message=FALSE, echo=FALSE, include=FALSE, purl=FALSE}
# p <- plotGraphSpace(gs, add.image = TRUE)
# ggsave(filename = "./vignettes/articles/cards/mapping_images.png", height=4, width=4,
#   units="in", device="png", dpi=200, plot = p)
```

## Orientation controls

The `normalizeGraphSpace()` function provides five arguments to adjust the alignment between graph coordinates and the background image. Arguments `flip.x`, `flip.y`, and `swap.xy` transform node coordinates; `flip.v` and `flip.h` transform the image matrix itself. The reference card below illustrates each argument using the `volcano` dataset, where nodes are precisely aligned with their corresponding dark pixels in the default configuration. Each panel shows how that alignment changes when a single argument is modified.

Note that `flip.y` defaults to `TRUE` when `image.space = TRUE`, since image matrices use a top-down row indexing that is the inverse of the standard graph coordinate system. The panel labelled `flip.y = FALSE` shows what happens when this default correction is suppressed.

```{r Orientation reference, eval=TRUE, message=FALSE, fig.width=12, fig.height=8, out.width="100%"}
library("patchwork")

# Helper: build and render one panel
make_panel <- function(..., title) {
  gs <- GraphSpace(gtoy3)
  gs_image(gs) <- as_colorraster(volcano2)
  gs <- normalizeGraphSpace(gs, image.space = TRUE, ...)
  plotGraphSpace(gs, add.image = TRUE) +
    ggplot2::labs(title = title) +
    ggplot2::theme(
      plot.title = element_text(
        size = 12, hjust = 0.5, face = "plain"),
      plot.margin = ggplot2::margin(1, 1, 10, 1)
    )
}

# One panel per orientation argument
p1 <- make_panel(title = "default")
p2 <- make_panel(flip.x  = TRUE,  title = "flip.x = TRUE")
p3 <- make_panel(flip.y  = FALSE, title = "flip.y = FALSE")
p4 <- make_panel(swap.xy = TRUE,  title = "swap.xy = TRUE")
p5 <- make_panel(flip.v  = TRUE,  title = "flip.v = TRUE")
p6 <- make_panel(flip.h  = TRUE,  title = "flip.h = TRUE")

# 3x2 reference grid
(p1 | p2 | p3) / (p4 | p5 | p6)
```

# Advanced workflows

See the [*Spatial Data*](spatial-data.html) tutorial for examples using a reference image.

# Session information
```{r label='Session information', eval=TRUE, echo=FALSE}
sessionInfo()
```

# References

