---
title: "Mapping Graphs to Images"
author: "Sysbiolab Team"
date: "`r Sys.Date()`"
bibliography: ../bibliography.bib
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
  )
```

<br/>
**Package**: RGraphSpace `r packageVersion('RGraphSpace')`

```{r Check required version, eval=TRUE, message=FALSE}
# Check required version
if (packageVersion("RGraphSpace") < "1.4.1"){
  message("Need to update 'RGraphSpace' for this vignette")
  remotes::install_github("sysbiolab/RGraphSpace")
}
```

# Overview

Images can be used as spatial references for graphs. When a raster image is provided, the pixel coordinates define where nodes are positioned, supporting the construction of graphs from image features.

# Setting basic input data

As an example, next we extract features from the `volcano` matrix and map them to graph nodes. The graph is then visualized over the image matrix.

```{r Mapping images - 1, eval=TRUE, message=FALSE}
library("RGraphSpace")
library("igraph")
library("ggplot2")

# Extract pixel coordinates for a specific intensity quantile.
coords <- which(volcano == quantile(volcano, 0.85), arr.ind = TRUE)

# Mark target pixels with '0'; it will appear as black in the background. 
# This creates a visual anchor to verify the alignment precision.
volcano2 <- volcano
volcano2[coords] <- 0

# Create an igraph object from the pixel coordinates; 
# note that at this stage, 'y' represents matrix row indices.
gtoy3 <- igraph::make_empty_graph(n = nrow(coords))
igraph::V(gtoy3)$y <- coords[,1]
igraph::V(gtoy3)$x <- coords[,2]

# Highlight the bottom-row vertex (max 'y' index) to demonstrate alignment; 
# since matrix indexing is top-down, this accounts for the default flip 
# between matrix and plot coordinate systems.
igraph::V(gtoy3)$nodeColor <- NA
bottom_row <- which.max(igraph::V(gtoy3)$y)
igraph::V(gtoy3)$nodeColor[bottom_row] <- adjustcolor("red", 0.4)
```

# Initialize, normalize, and plot

```{r Mapping images - 2, eval=TRUE, message=FALSE}
# Initialize a GraphSpace object
gs <- GraphSpace(gtoy3)
gs_image(gs) <- as_colorraster(volcano2)

# Map graph coordinates to the image space; by default,
# 'y' row indices will be flipped (see comments below).
gs <- normalizeGraphSpace(gs)
```

Observe that the nodes (open circles) are precisely aligned with the pixels (dark squares) to which they were mapped. In this plot, nodes remain proportional to the viewport regardless of the final output resolution. This behavior is especially critical when the network elements represent image features at specific positions and must stay anchored to those features, avoiding overlap with adjacent, unrelated areas.

```{r Mapping images - 3, eval=TRUE, message=FALSE}
# Render the graph with the raster as background
plotGraphSpace(gs, add.image = TRUE)
```

**Note on image alignment**: Proper spatial alignment between nodes and the background image requires consistent coordinate conventions. Spatial misalignment may occur if the input image and node coordinates differ in axis orientation (e.g., top-left versus bottom-left origins). To accommodate these differences, `normalizeGraphSpace()` provides orientation controls through the `rotate.xy`, `flip.x`, and `flip.y` arguments. If the nodes appear misaligned with the input image, try combinations of these parameters to correct the alignment. Alternatively, try `flip.v` and `flip.h` arguments to apply flipping directly to the background image.

```{r Mapping images - 4, eval=TRUE, message=FALSE}
# Alternatively, passing to the underlying geoms
ggplot(gs) +
  annotation_gspace_image(gs) +
  geom_nodespace() +
  theme_gspace_coords(is_norm = TRUE)
```

# Advanced workflows

See *PathwaySpace* tutorials for use-case examples involving reference image backgrounds.

* <a href="https://sysbiolab.github.io/PathwaySpace/" target="_blank">https://sysbiolab.github.io/PathwaySpace/</a>


```{r, eval=FALSE, message=FALSE, echo=FALSE, include=FALSE, purl=FALSE}
# p <- plotGraphSpace(gs, add.image = TRUE)
# ggsave(filename = "./vignettes/cards/mapping_images.png", height=4, width=4,
#   units="in", device="png", dpi=200, plot = p)
```


# Session information
```{r label='Session information', eval=TRUE, echo=FALSE}
sessionInfo()
```


